taxize - taxonomic search and retrieval in R

### 2013-06-10 inst/doc/taxize_withcode_app.tex
more work on example

First we look if we get any direct matches between taxon names:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# first search for direct matches}
direct <- \hlfunctioncall{match}(lookup$taxon, traits$taxon_cleaned)
\hlcomment{# and add the matched name to our lookup table}
lookup$traits <- \hlfunctioncall{tolower}(traits$taxon[direct])
lookup$type <- \hlfunctioncall{ifelse}(!\hlfunctioncall{is.na}(direct), \hlstring{"direct"}, NA)
lookup
\end{alltt}
\begin{verbatim}
                     taxon                   traits   type
1         gammarus roeseli                     <NA>   <NA>
2        gammarus tigrinus                     <NA>   <NA>
3           coenagrionidae                     <NA>   <NA>
4 potamopyrgus antipodarum potamopyrgus antipodarum direct
5                    xxxxx                     <NA>   <NA>
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

Next we look for species which are on a better taxonomic resolution than our in our trait table. 
For these species we'll use the trait-data since no better is available.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# look for cases where taxonomic resolution in abundance data is higher}
\hlcomment{# than in trait data: here we take the trait-values for the lower}
\hlcomment{# resolution}
\hlfunctioncall{for} (i in \hlfunctioncall{which}(\hlfunctioncall{is.na}(lookup$traits))) \{
    \hlfunctioncall{if} (\hlfunctioncall{is.data.frame}(lookup_classi[[i]])) \{
        matches <- \hlfunctioncall{tolower}(lookup_classi[[i]]$ScientificName) %in% traits$taxon_cleaned
        \hlfunctioncall{if} (\hlfunctioncall{any}(matches)) \{
            lookup$traits[i] <- \hlfunctioncall{tolower}(lookup_classi[[i]]$ScientificName[matches])
            lookup$type[i] <- lookup_classi[[i]]$Rank[matches]
        \}
    \}
\}
lookup
\end{alltt}
\begin{verbatim}
                     taxon                   traits   type
1         gammarus roeseli                 gammarus  genus
2        gammarus tigrinus                 gammarus  genus
3           coenagrionidae                     <NA>   <NA>
4 potamopyrgus antipodarum potamopyrgus antipodarum direct
5                    xxxxx                     <NA>   <NA>
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

Our next step is to search for species were we have to aggregate tait-data, since our abundance data is on a lower taxonomic level.
We are walking the taxomomic latter for the species in our trait-data upwards and search for matches with out abundance data. Since we'll have many taxa in the trait-data belonging to one taxon, we'll take the median modality scores as an approximation. Of course also other methods may be used here, e.g. weighting by genetic distance.

@@

\hlfunctioncall{for} (i in \hlfunctioncall{which}(\hlfunctioncall{is.na}(lookup$traits))) \{
\hlcomment{    # find matches}
    agg <- \hlfunctioncall{sapply}(traits_classi, \hlfunctioncall{function}(x) \hlfunctioncall{any}(\hlfunctioncall{tolower}(x$ScientificName) %in% 
        lookup$taxon[i]))
    \hlfunctioncall{if} (\hlfunctioncall{sum}(agg) > 1) \{
\hlcomment{        # add taxon as aggregate to trait-table}
        traits <- \hlfunctioncall{rbind}(traits, \hlfunctioncall{c}(\hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"}), \hlfunctioncall{apply}(traits[agg, 
            2:6], 2, median), \hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"})))
\hlcomment{        # fill lookup table}
        lookup$traits[i] <- \hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"})
        lookup$type[i] <- \hlstring{"aggregated"}
    \}
\}
lookup
\end{alltt}
\begin{verbatim}
                     taxon                    traits       type
1         gammarus roeseli                  gammarus      genus
2        gammarus tigrinus                  gammarus      genus
3           coenagrionidae coenagrionidae-aggregated aggregated
4 potamopyrgus antipodarum  potamopyrgus antipodarum     direct
5                    xxxxx                      <NA>       <NA>
\end{verbatim}
\end{kframe}
\end{knitrout}

Finally we have one taxon left. We remove this from our dataset:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
abundances <- abundances[!abundances$taxon == lookup$taxon[\hlfunctioncall{is.na}(lookup$traits)], 
    ]
\end{alltt}
\end{kframe}
\end{knitrout}

No we can create species x sites and traits x species matrices, which could be plugged into methods to analyse trait responses \citep{kleyer_assessing_2012}.

@@

>>>>

First we look if we can find any direct matches between taxon names:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# first search for direct matches}
direct <- \hlfunctioncall{match}(lookup$taxon, traits$taxon_cleaned)
\hlcomment{# and add the matched name to our lookup table}
lookup$traits <- \hlfunctioncall{tolower}(traits$taxon[direct])
lookup$match <- \hlfunctioncall{ifelse}(!\hlfunctioncall{is.na}(direct), \hlstring{"direct"}, NA)
lookup
\end{alltt}
\begin{verbatim}
                     taxon                   traits  match
1         gammarus roeseli                     <NA>   <NA>
2        gammarus tigrinus                     <NA>   <NA>
3           coenagrionidae                     <NA>   <NA>
4 potamopyrgus antipodarum potamopyrgus antipodarum direct
5                    xxxxx                     <NA>   <NA>
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

Next we look for species which are on a higher taxonomic resolution than our trait table. 
For these species we will take directly the trait-data since no better information is available.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcomment{# look for cases where taxonomic resolution in abundance data is higher}
\hlcomment{# than in trait data: here we take the trait-values for the lower}
\hlcomment{# resolution}
\hlfunctioncall{for} (i in \hlfunctioncall{which}(\hlfunctioncall{is.na}(lookup$traits))) \{
    \hlfunctioncall{if} (\hlfunctioncall{is.data.frame}(lookup_classi[[i]])) \{
        matches <- \hlfunctioncall{tolower}(lookup_classi[[i]]$ScientificName) %in% traits$taxon_cleaned
        \hlfunctioncall{if} (\hlfunctioncall{any}(matches)) \{
            lookup$traits[i] <- \hlfunctioncall{tolower}(lookup_classi[[i]]$ScientificName[matches])
            lookup$match[i] <- lookup_classi[[i]]$Rank[matches]
        \}
    \}
\}
lookup
\end{alltt}
\begin{verbatim}
                     taxon                   traits  match
1         gammarus roeseli                 gammarus  genus
2        gammarus tigrinus                 gammarus  genus
3           coenagrionidae                     <NA>   <NA>
4 potamopyrgus antipodarum potamopyrgus antipodarum direct
5                    xxxxx                     <NA>   <NA>
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

The next step is to search for species were we have to aggregate tait-data, since our abundance data is on a lower taxonomic level.
We are walking the taxomomic latter for the species in our trait-data upwards and search for matches with out abundance data. Since we'll have many taxa in the trait-data belonging to one taxon, we'll take the median modality scores as an approximation. Of course also other methods may be used here, e.g. weighting by genetic distance.

@@

\hlfunctioncall{for} (i in \hlfunctioncall{which}(\hlfunctioncall{is.na}(lookup$traits))) \{
\hlcomment{    # find matches}
    agg <- \hlfunctioncall{sapply}(traits_classi, \hlfunctioncall{function}(x) \hlfunctioncall{any}(\hlfunctioncall{tolower}(x$ScientificName) %in% 
        lookup$taxon[i]))
    \hlfunctioncall{if} (\hlfunctioncall{sum}(agg) > 1) \{
\hlcomment{        # add taxon as aggregate to trait-table}
        traits <- \hlfunctioncall{rbind}(traits, \hlfunctioncall{c}(\hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"}), \hlfunctioncall{apply}(traits[agg, 
            2:6], 2, median), \hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"})))
\hlcomment{        # fill lookup table}
        lookup$traits[i] <- \hlfunctioncall{paste0}(lookup$taxon[i], \hlstring{"-aggregated"})
        lookup$match[i] <- \hlstring{"aggregated"}
    \}
\}
lookup
\end{alltt}
\begin{verbatim}
                     taxon                    traits      match
1         gammarus roeseli                  gammarus      genus
2        gammarus tigrinus                  gammarus      genus
3           coenagrionidae coenagrionidae-aggregated aggregated
4 potamopyrgus antipodarum  potamopyrgus antipodarum     direct
5                    xxxxx                      <NA>       <NA>
\end{verbatim}
\end{kframe}
\end{knitrout}

Finally we have only one taxon left - clearly an error. We remove this from our dataset:
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
abundances <- abundances[!abundances$taxon == lookup$taxon[\hlfunctioncall{is.na}(lookup$traits)], 
    ]
\end{alltt}
\end{kframe}
\end{knitrout}

No we can create \emph{species x sites} and \emph{traits x species} matrices, which could be plugged into methods to analyse trait responses \citep{kleyer_assessing_2012}.

@@

This is just an example how taxonomic APIs (via taxize) could be used to search for matches up- and downwards the taxonomic ladder. We are looking forward to integrate the freshwaterecology.info database \citep{freshwaterecology} into taxize, which will facilitate trait-based analyses in R.

### 2013-06-10 inst/doc/taxize_withcode_appB.Rnw
more work on example

Suppose we have fuzzy coded trait table with 2 traits with 3 respectively 2 modalities:
<<traitdata>>=
(traits <- read.table(header = TRUE, sep = ';', stringsAsFactors=FALSE, text = 'taxon;T1M1;T1M2;T1M3;T2M1;T2M2
Gammarus sp.;0;0;3;1;3
Potamopyrgus antipodarum;1;0;3;1;3
Coenagrion sp.;3;0;1;3;1
Enallagma cyathigerum;0;3;1;0;3
Erythromma sp.;0;0;3;3;1
Baetis sp.;0;0;0;0;0
'))
@

And want to match this to a table with abundances:
<<abundancedata>>=
(abundances <- read.table(header = TRUE, sep = ';', stringsAsFactors=FALSE, text = 'taxon;abundance;sample
Gammarus roeseli;5;1
Gammarus roeseli;6;2
Gammarus tigrinus;7;1
Gammarus tigrinus;8;2
Coenagrionidae;10;1
Coenagrionidae;6;2
Potamopyrgus antipodarum;10;1
xxxxx;10;2
'))
@

First we do some basic data-cleaning and create a looktable, that will link taxa in trait table with taxa in the abundance table.
<<cleaning>>=
# first we remove ' sp.' from out trait table:
traits$taxon_cleaned <- tolower(gsub(' sp.', '', traits$taxon))

@@

The we query the taxonomic hierarchy for both tables, since this is the backbone of this procedure:
<<get_classification, cache=TRUE>>=
require(taxize)
traits_classi <- classification(get_uid(traits$taxon_cleaned))
lookup_classi <- classification(get_uid(lookup$taxon))
@

First we look if we get any direct matches between taxon names:
<<direct_matches>>=
# first search for direct matches
direct <- match(lookup$taxon, traits$taxon_cleaned)
# and add the matched name to our lookup table
lookup$traits <- tolower(traits$taxon[direct])
lookup$type <- ifelse(!is.na(direct), "direct", NA)
lookup
@

@@

Next we look for species which are on a better taxonomic resolution than our in our trait table. 
For these species we'll use the trait-data since no better is available.

<<higher_resolution>>=
# look for cases where taxonomic resolution in abundance data is higher than in trait data:
# here we take the trait-values for the lower resolution
for(i in which(is.na(lookup$traits))){
  if(is.data.frame(lookup_classi[[i]])){
    matches <- tolower(lookup_classi[[i]]$ScientificName) %in% traits$taxon_cleaned
    if(any(matches)){
      lookup$traits[i] <- tolower(lookup_classi[[i]]$ScientificName[matches])
      lookup$type[i] <- lookup_classi[[i]]$Rank[matches]
    }
  }
}
lookup
@

So our abundance data has two Gammarus species, however trait data is only on Genus level.

Our next step is to search for species were we have to aggregate tait-data, since our abundance data is on a lower taxonomic level.
We are walking the taxomomic latter for the species in our trait-data upwards and search for matches with out abundance data. Since we'll have many taxa in the trait-data belonging to one taxon, we'll take the median modality scores as an approximation. Of course also other methods may be used here, e.g. weighting by genetic distance.

@@

for(i in which(is.na(lookup$traits))){
  # find matches
  agg <- sapply(traits_classi, function(x) any(tolower(x$ScientificName) %in% lookup$taxon[i]))
  if(sum(agg) > 1){
    # add taxon as aggregate to trait-table
    traits <- rbind(traits, c(paste0(lookup$taxon[i], '-aggregated'), apply(traits[agg, 2:6], 2, median), paste0(lookup$taxon[i], '-aggregated')))
    # fill lookup table
    lookup$traits[i] <- paste0(lookup$taxon[i], '-aggregated')
    lookup$type[i] <- 'aggregated'
  }
}
lookup
@

Finally we have one taxon left. We remove this from our dataset:
<<na_omit>>=
abundances <- abundances[!abundances$taxon == lookup$taxon[is.na(lookup$traits)], ]
@

No we can create species x sites and traits x species matrices, which could be plugged into methods to analyse trait responses \citep{kleyer_assessing_2012}.

>>>>

Suppose we have fuzzy coded trait table with 2 traits with 3 respectively 2 modalities:
<<traitdata, tidy=FALSE>>=
(traits <- read.table(header = TRUE, sep = ';', stringsAsFactors=FALSE, 
                      text = 'taxon;T1M1;T1M2;T1M3;T2M1;T2M2
Gammarus sp.;0;0;3;1;3
Potamopyrgus antipodarum;1;0;3;1;3
Coenagrion sp.;3;0;1;3;1
Enallagma cyathigerum;0;3;1;0;3
Erythromma sp.;0;0;3;3;1
Baetis sp.;0;0;0;0;0
'))
@

And want to match this to a table with abundances:
<<abundancedata, tidy=FALSE>>=
(abundances <- read.table(header = TRUE, sep = ';', stringsAsFactors=FALSE, 
                          text = 'taxon;abundance;sample
Gammarus roeseli;5;1
Gammarus roeseli;6;2
Gammarus tigrinus;7;1
Gammarus tigrinus;8;2
Coenagrionidae;10;1
Coenagrionidae;6;2
Potamopyrgus antipodarum;10;1
xxxxx;10;2
'))
@

First we do some basic data-cleaning and create a lookup-table, that will link taxa in trait table with the abundance table.
<<cleaning>>=
# first we remove ' sp.' from out trait table:
traits$taxon_cleaned <- tolower(gsub(' sp.', '', traits$taxon))

@@

The we query the taxonomic hierarchy for both tables, this will be the backbone of this procedure:
<<get_classification, cache=TRUE>>=
require(taxize)
traits_classi <- classification(get_uid(traits$taxon_cleaned))
lookup_classi <- classification(get_uid(lookup$taxon))
@

First we look if we can find any direct matches between taxon names:
<<direct_matches>>=
# first search for direct matches
direct <- match(lookup$taxon, traits$taxon_cleaned)
# and add the matched name to our lookup table
lookup$traits <- tolower(traits$taxon[direct])
lookup$match <- ifelse(!is.na(direct), "direct", NA)
lookup
@

@@

Next we look for species which are on a higher taxonomic resolution than our trait table. 
For these species we will take directly the trait-data since no better information is available.

<<higher_resolution>>=
# look for cases where taxonomic resolution in abundance data is higher than in trait data:
# here we take the trait-values for the lower resolution
for(i in which(is.na(lookup$traits))){
  if(is.data.frame(lookup_classi[[i]])){
    matches <- tolower(lookup_classi[[i]]$ScientificName) %in% traits$taxon_cleaned
    if(any(matches)){
      lookup$traits[i] <- tolower(lookup_classi[[i]]$ScientificName[matches])
      lookup$match[i] <- lookup_classi[[i]]$Rank[matches]
    }
  }
}
lookup
@

So our abundance data has two Gammarus species, however trait data is only on genus level.

The next step is to search for species were we have to aggregate tait-data, since our abundance data is on a lower taxonomic level.
We are walking the taxomomic latter for the species in our trait-data upwards and search for matches with out abundance data. Since we'll have many taxa in the trait-data belonging to one taxon, we'll take the median modality scores as an approximation. Of course also other methods may be used here, e.g. weighting by genetic distance.

@@

for(i in which(is.na(lookup$traits))){
  # find matches
  agg <- sapply(traits_classi, function(x) any(tolower(x$ScientificName) %in% lookup$taxon[i]))
  if(sum(agg) > 1){
    # add taxon as aggregate to trait-table
    traits <- rbind(traits, c(paste0(lookup$taxon[i], '-aggregated'), apply(traits[agg, 2:6], 2, median), paste0(lookup$taxon[i], '-aggregated')))
    # fill lookup table
    lookup$traits[i] <- paste0(lookup$taxon[i], '-aggregated')
    lookup$match[i] <- 'aggregated'
  }
}
lookup
@

Finally we have only one taxon left - clearly an error. We remove this from our dataset:
<<na_omit>>=
abundances <- abundances[!abundances$taxon == lookup$taxon[is.na(lookup$traits)], ]
@

No we can create \emph{species x sites} and \emph{traits x species} matrices, which could be plugged into methods to analyse trait responses \citep{kleyer_assessing_2012}.
